---
title: "CSP571_FIFA_PLAYER_PREDICTION_ANALYSIS"
output: pdf_document
date: "2024-04-21"
---

#### Importing Libraries:

```{r}
library(ggplot2)
library(dplyr)
library(gridExtra)
```

#### Loading the Data Set

```{r}
df = read.csv(file = "./CompleteDataset.csv", stringsAsFactors = FALSE)
df <- select(df, ID, X, Name, Age, Nationality, Overall, Club, Value, Wage, Preferred.Positions)

head(df, 10)

```

#### Data Cleaning

The first thing that we did after reading in the dataset was to convert the `Value` and the `Wage` columns to actual currency values. We wrote a function that took a vector as an input and removed the "€" sign from the columns and multiplied it with appropriate number to convert it into thousand(K) and million(M).

```{r}
toNumberCurrency <- function(vector) {
    vector <- as.character(vector)
    vector <- gsub("(€|,)","", vector)
    result <- suppressWarnings(as.numeric(vector))
   
    k_positions <- grep("K", vector)
    result[k_positions] <- as.numeric(gsub("K","",vector[k_positions])) * 1000
  
    m_positions <- grep("M", vector)
    result[m_positions] <- as.numeric(gsub("M","", 
    vector[m_positions])) * 1000000
    
    return(result)
}
df$Wage <- toNumberCurrency(df$Wage) 
df$Value <- toNumberCurrency(df$Value)
```

Then we picked the first position-value from the space separated values in the `Preferred.Positions` column. The different positions in the dataset were: `CAM CB CDM CF CM GK LB LM LW LWB RB RM RW RWB ST` .

```{r}
df$Preferred.Positions <- gsub(" ", "", substr(df$Preferred.Positions, 1, 3))
```

Based on the above `Preferred.Positions` , We created another column that classified these `Preferred.Positions` into more general positions namely, `GK DEF MID FWD` .

#### Data Transformations

Select the first position in the preferred.Position column and create a table to classify these into more general positions namely GK, DEF MID FWD.

```{r}
x <- as.factor(df$Preferred.Positions)
levels(x) <- list(GK  = c("GK"), 
                  DEF = c("LWB", "LB", "CB", "RB", "RWB"), 
                  MID = c("LW","LM","CDM","CM","CAM","RM","RW"), 
                  FWD = c("CF", "ST"))
df <- mutate(df, Position = x)
head(df)
```

### **Plotting Distribution of Player Ages**

Firstly, distribution of players based on the age. We see that there is a high number of players around 25 years of age.

```{r}
g_age <- ggplot(data = df, aes(Age))
g_age + 
geom_histogram(col="orange", aes(fill = ..count..)) + ggtitle("Distribution based on Age")
```

### **Visualizing Age Distribution by Player Position**

The following plot shows the relation between the Age of the players and their general playing position.

```{r}
g_age + 
geom_density(col="orange", aes(fill = Position), alpha=0.5) + facet_grid(.~Position) + 
ggtitle("Distribution based on Age and Position")
```

#### **Visualizing Distribution of Overall Ratings**

```{r}
g_overall <- ggplot(data = df, aes(Overall))
g_overall + 
geom_histogram(col="orange", aes(fill = ..count..)) + ggtitle("Distribution based on Overall Rating")
```

#### **Visualizing Distribution of Players by Nationality (Top 10 Countries)**

```{r}
countries_count <- count(df, Nationality)
top_10_countries <- top_n(countries_count, 10, n)
top_10_country_names <- top_10_countries$Nationality

country <- filter(df, Nationality == top_10_country_names)
ggplot(country, aes(x = Nationality)) + 
geom_bar(col = "orange", aes(fill = ..count..)) + ggtitle("Distribution based on Nationality of Players (Top 10 Countries)")
```

### **Visualizing Distribution of Top 1% Wage**

We used the quantile function in R to obtain the top 1 % count of the player `Value` and `Wage` . A large number of players earn a weekly wage of €100000.

```{r}
top_1_percent_wage   <- quantile(df$Wage, probs=0.99)
filtered_wage <- filter(df, Wage > top_1_percent_wage)

g_value <- ggplot(filtered_wage, aes(Wage))
g_value + 
geom_histogram(aes(fill=..count..)) + 
ggtitle("Distribution of top 1% value")
```

#### **Visualizing Distribution of Top 1% Player Values**

A large number of players are worth €25,000,000 with the count decreasing sharply as the price increases.

```{r}
top_1_percent_value   <- quantile(df$Value, probs=0.99)
filtered_value <- filter(df, Value > top_1_percent_value)

g_wage <- ggplot(filtered_value, aes(Value))
g_wage + 
geom_histogram(aes(fill=..count..)) + 
ggtitle("Distribution of top 1% Value")
```

### **Creating Wage and Value Brackets for Player Data**

To further analyze the `wage` and `value` , I created `wage_brackets` and `value_brackets` .

The `wage_brackets` are: `0–100k 100k-200k 200k-300k 300k-400k 400k-500k 500k+` . The `value_brackets` are : `0–10M 10–20M 20–30M 30–40M 40–50M 50–60M 60–70M 70–80M 80–90M 90–100M 100M+` .

```{r}
# Define wage breaks and labels for creating wage brackets
wage_breaks <- c(0, 100000, 200000, 300000, 400000, 500000, Inf)
wage_labels <- c("0-100k", "100k-200k", "200k-300k", "300k-400k", "400k-500k", "500k+")

# Create wage brackets based on defined breaks and labels
wage_brackets <- cut(x = df$Wage, breaks = wage_breaks, labels = wage_labels, include.lowest = TRUE)

# Add wage brackets as a new column 'wage_brackets' to the dataframe
df <- mutate(df, wage_brackets)

# Define value breaks and labels for creating value brackets
value_breaks <- c(0, 10000000, 20000000, 30000000, 40000000, 50000000, 60000000, 70000000, 80000000, 90000000, 100000000, Inf)
value_labels <- c("0-10M", "10-20M", "20-30M", "30-40M", "40-50M", "50-60M", "60-70M", "70-80M", "80-90M", "90-100M", "100M+")

# Create value brackets based on defined breaks and labels
value_brackets <- cut(x = df$Value, breaks = value_breaks, labels = value_labels, include.lowest = TRUE)

# Add value brackets as a new column 'value_brackets' to the dataframe
df <- mutate(df, value_brackets)

# Display the first few rows of the updated dataframe
head(df)

```

#### **Visualizing Distribution of Top Wage between 100K-500K+**

A very large number of players have wages which lie between 0--100k and valuation between 0--50M . Plotting these values skewed the graphs a lot since they are high in magnitude as compared to the rest of the values, hence I haven't shown them here.

```{r}
not0To100K <- filter(df, wage_brackets != "0-100k") 
ggplot(not0To100K, aes(x = wage_brackets)) + 
geom_bar(aes(fill = ..count..)) + 
ggtitle("Distribution of top Wage between 100K-500K+")
```

### **Visualizing Distribution of Value between 50M-100M+**

```{r}
moreThan50M <- filter(df, Value>50000000)
ggplot(moreThan50M, aes(x = value_brackets)) + 
geom_bar(aes(fill = ..count..)) + 
ggtitle("Distribution of value between 50M-100M+")
```

#### **Exploring Relationship between Age, Overall, and Wage Brackets**

Age vs Overall of players divided amongst wage brackets. The highest wages are commanded by players of overall 85+ and age around 30 years.

```{r}
g_age_overall <- ggplot(df, aes(Age, Overall))
g_age_overall + 
geom_point(aes(color=wage_brackets)) + geom_smooth(color="darkblue") + 
ggtitle("Distribution between Age and Overall of players based  on Wage bracket")
```

#### **Exploring Relationship between Age, Overall, and Value Brackets**

Number of players as per their general playing positions. Number of midfielders is the highest followed by defenders, forwards, and finally goalkeepers.

```{r}
g_age_overall <- ggplot(df, aes(Age, Overall))
g_age_overall + geom_point(aes(color=value_brackets)) + geom_smooth(color="darkblue") + 
ggtitle("Distribution between Age and Overall of players based on Value bracket")
```

#### **Visualizing Distribution of General Playing Positions**

```{r}
ggplot(df, aes(Position)) + 
geom_bar(aes(fill = ..count..)) + 
ggtitle("Distribution based on General Playing Position")
```

### **Visualizing Distribution of Players based on Preferred Position**

Number of players as per their preferred playing positions. Based on the above graph, we'd expect some specific midfielder position to have the highest count, but here number of center-backs is the highest followed by the number of strikers.

```{r}
ggplot(df, aes(Preferred.Positions)) + geom_bar(aes(fill=..count..)) + 
ggtitle("Distribution of players based on preferred position")
```

### **Visualizing Position Distribution based on Player Value (0-50M and 50M+ Categories)**

```{r}
gf1 <- filter(df, Value<30000000)
g1 <- ggplot(gf1, aes(Preferred.Positions)) + geom_bar(aes(fill=value_brackets)) + 
ggtitle("Position based on Value (0-50M)")
gf2 <- filter(df,Value>30000000)
g2 <- ggplot(gf2, aes(Preferred.Positions)) + geom_bar(aes(fill=value_brackets)) + 
ggtitle("Position based on Value (50M +)")
grid.arrange(g1, g2, ncol=1)
```

### **Visualizing Position Distribution based on Player Wage (0-100k and 100k+ Categories)**

```{r}
gw1 <- filter(df, Wage > 100000, Wage<300000)
g1 <- ggplot(gw1, aes(Preferred.Positions)) + geom_bar(aes(fill=wage_brackets)) + 
ggtitle("Position based on Wage (0-100k)") 
gw2 <- filter(df,Wage>300000) 
g2 <- ggplot(gw2, aes(Preferred.Positions)) + geom_bar(aes(fill=wage_brackets)) + 
ggtitle("Position based on Wage (100k+)")
grid.arrange(g1, g2, ncol=1)
```

#### **Visualizing Top 10 Valuable Clubs**

The top ten valuable clubs. The club value is calculated by summing up the player valuation for each club.

```{r}
group_clubs <- group_by(df, Club)
club_value <- summarise(group_clubs, total_val = sum(Value))
top_10_valuable_clubs <- top_n(club_value, 10, total_val)

top_10_valuable_clubs$Club <-as.factor(top_10_valuable_clubs$Club)

ggplot(top_10_valuable_clubs, aes(x = Club, y = total_val)) + geom_bar(stat = "identity", aes(fill=total_val)) + coord_flip() + ggtitle("Top 10 valuable clubs")
```

Importing Libraries for processing

```{r}
library(tidyverse)  # for data wrangling
library(stringr)    # for string manipulations
library(ggbiplot)   # pca biplot with ggplot
library(Rtsne)      # implements the t-SNE algorithm
library(kohonen)    # implements self organizing maps
library(hrbrthemes) # nice themes for ggplot
library(GGally)     # to produce scatterplot matrices
library(dplyr)
```

```{r}
fifa_tbl = read.csv(file = "./CompleteDataset.csv", stringsAsFactors = FALSE)
names(fifa_tbl)
```

#### Model Training

```{r}
library(dplyr)
# Selecting columns explicitly from 'Acceleration' to 'Volleys' and preferred positions.
fifa_tbl <- fifa_tbl %>%
  dplyr::select(
    `Acceleration`, 
    `Aggression`, 
    `Agility`, 
    `Balance`, 
    `Ball.control`, 
    `Composure`, 
    `Crossing`, 
    `Curve`, 
    `Dribbling`, 
    `Finishing`, 
    `Free.kick.accuracy`, 
    `GK.diving`, 
    `GK.handling`, 
    `GK.kicking`, 
    `GK.positioning`, 
    `GK.reflexes`, 
    `Heading.accuracy`, 
    `Interceptions`, 
    `Jumping`, 
    `Long.passing`, 
    `Long.shots`, 
    `Marking`, 
    `Penalties`, 
    `Positioning`, 
    `Reactions`, 
    `Short.passing`, 
    `Shot.power`, 
    `Sliding.tackle`, 
    `Sprint.speed`, 
    `Stamina`, 
    `Standing.tackle`, 
    `Strength`, 
    `Vision`, 
    `Volleys`, 
    `Preferred.Positions` 
  )
```

### **Transforming Player Positions in FIFA Data**

```{r}
fifa_tbl <- fifa_tbl %>% 
  mutate(position = word(`Preferred.Positions`,1)) %>% 
  mutate(position = factor(position,
                           levels = c("GK","CB","RB","LB","RWB","LWB","CDM",
                                      "CM","RM","LM","CAM",
                                      "CF","RW","LW","ST")))

defense  <- c("CB","RB","LB","RWB","LWB")
midfield <- c("CDM","CM","RM","LM","CAM")
offense  <- c("CF","RW","LW","ST")
    
fifa_tbl <- fifa_tbl %>% 
  mutate(position2 = ifelse(position %in% defense,"D",
                     ifelse(position %in% midfield,"M",
                     ifelse(position %in% offense,"O","GK")))) %>% 
  mutate(position2 = factor(position2,levels = c("GK","D","M","O"))) %>% 
  dplyr::select(-`Preferred.Positions`)
```

Importing ML library

```{r}
library(caret) # ML algorithms
```

```{r}
glimpse(fifa_tbl)
```

```{r}
all_levels <- unique(fifa_tbl$Acceleration)
fifa_tbl$Acceleration <- factor(fifa_tbl$Acceleration, levels = all_levels)
```

#### Splitting the Data in Training and Testing

```{r}
#### Splitting the Data in Training and Testing
set.seed(6185)

categorical_columns <- sapply(fifa_tbl, is.factor)  # Identify factor columns
levels_list <- lapply(fifa_tbl[, categorical_columns], unique)  # List of all levels for each factor

# Apply all levels to each factor in the dataset
for (col in names(levels_list)) {
  levels_set <- unique(unlist(levels_list[col]))
  fifa_tbl[[col]] <- factor(fifa_tbl[[col]], levels = levels_set)
}

# Create a smaller subset of the data
small_data <- fifa_tbl[sample(nrow(fifa_tbl), size = 1000), ]

# Split the smaller data into train and test sets
train_sample <- createDataPartition(small_data$position2, p = 0.8, list = FALSE)
train_data <- small_data[train_sample, ] %>% select(-position)
test_data <- small_data[-train_sample, ] %>% select(-position)

train_data$Acceleration <- factor(train_data$Acceleration, levels = all_levels)
test_data$Acceleration <- factor(test_data$Acceleration, levels = all_levels)
```

```{r}
library(doParallel)  # Load the parallel processing library
registerDoParallel(cores = detectCores() - 1)  # Register the number of cores (leave one out for system stability)
```

### KNN:

```{r}
# Define train_control object
train_control <- trainControl(method = "repeatedcv",
                              number = 10,
                              repeats = 3,
                              savePredictions = TRUE,
                              classProbs = TRUE)
```

```{r}
# Train KNN on the sampled data
fifa_knn <- suppressWarnings(train(position2 ~ ., data = train_data, method = "knn",
                  trControl = train_control, preProcess = c("center", "scale"),
                  tuneGrid = expand.grid(.k = seq(20, 120, 5))))
fifa_knn
```

```{r}
plot(fifa_knn)
```

```{r}
#fifa_knn_predict <- predict(fifa_knn,new data = test_data)
fifa_knn_predict <- predict(fifa_knn, newdata = test_data)
confusionMatrix(fifa_knn_predict,test_data$position2)
```

### Random Forrest

```{r}
# Enable parallel processing and reduce the number of trees for quicker initial results
train_control_rf <- trainControl(method = "repeatedcv",
                                 number = 5,
                                 repeats = 1,
                                 allowParallel = TRUE,
                                 savePredictions = "final")
grid_rf <- expand.grid(.mtry = c(4, 8, 12, 16, 20))  # Tuning mtry only, adjust based on feature count
fifa_rf <- suppressWarnings(train(position2 ~ ., data = train_data, method = "rf",
                 trControl = train_control_rf, preProcess = c("center", "scale"),
                 ntree = 100,  # Reduced from the default 100 for faster preliminary results
                 tuneGrid = grid_rf))
fifa_rf
```

```{r}
plot(fifa_rf)
```

```{r}
fifa_rf_predict <- predict(fifa_rf, newdata = test_data)
confusionMatrix(fifa_rf_predict, test_data$position2)
```

### Support Vector Machine (SVM)

```{r}
# Use a narrower and less granular range for the cost parameter
grid_svm <- expand.grid(.cost = c(0.75, 0.9, 1, 1.1, 1.25,2))
fifa_svm_linear <- suppressWarnings(
  train(position2 ~ ., data = train_data, method = "svmLinear2",
        trControl = train_control, preProcess = c("center", "scale"),
        tuneGrid = grid_svm)
)
fifa_svm_linear
```

```{r}
plot(fifa_svm_linear)
```

```{r}
fifa_svm_linear_predict <- predict(fifa_svm_linear, newdata = test_data)
confusionMatrix(fifa_svm_linear_predict, test_data$position2)
```

```{r}
#best training performances
train_performance <- t(rbind(knn = apply(fifa_knn$results[, 2:3], 2, max),
                             rf  = apply(fifa_rf$results[, 2:3], 2, max),
                             svm = apply(fifa_svm_linear$results[, 2:3], 2, max)))

predicted <- tibble(knn = fifa_knn_predict,
                    rf  = fifa_rf_predict,
                    svm = fifa_svm_linear_predict)

#test performance
test_performance <- apply(predicted, 2, postResample, obs = test_data$position2)
```
